---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ofemeantest

<!-- badges: start -->
<!-- badges: end -->

## Installation

You can install the development version of ofemeantest from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pkg_install("PPaccioretti/ofemeantest")
```

## Example

The example dataset comprised a field trial of corn (Zea mays L.), where the
impact of broadcast phosphorus fertilization was evaluated. The treatment strip
received superphosphate at rates ranging from 250 to 500 $kg~ha^{–1}$ [0-21-0],
while the control strip received no P fertilizer (0 $kg~ha^{–1}$).
Each strip was 2.2 ha in area, totaling 100 ha. Raw grain yield monitor data 
for each each single strip were cleaned following the protocol proposed by 
Vega et al. (2019).


```{r example}
library(ofemeantest)
data("ofe_f2")
```

The `ofemt` function first creates a regular artificial grid over the area, 
including the strips to be analyzed. The grid spacing (`cellsize`) was set 
to 9 m. This grid overlaid the point data from the yield map, assigning
each georeferenced data point to a corresponding grid cell. Each grid cell 
contained a minimum of four data points (`nmin_cell = 4`). Empty cells or cells
with less than five data points are removed. The median of the yield
data is calculated for each cell. 
A one-way ANOVA is fitted to the cell medians to compare treatments.
From the residuals of the fitted model, the magnitude of spatial 
autocorrelation is estimated using the approximate profile likelihood 
estimator of spatial dependence (Rho) and the Moran index (MI). An effective
sample size (ESS) is then calculated.
Second, `ofemt` randomly samples grid cells without replacement using the ESS.
The yield data from both strips (treatment and control) are used to perform 
a permutational ANOVA with 2000 iterations per test (`n_p`).
This procedure provides a p-value for the comparison of treatment means. 
The permutational ANOVA is repeated on 200 random samples of grid cells (`n_s`)
using the same sampling procedure. The significance level for comparing means
(`alpha`), is left at their default value (0.05). The resulting permutation
p-values are collected to generate an empirical p-value distribution. 
Finally, the median p-value is calculated to compare the means.

```{r use_function}
ofemt(data = ofe_f2,
      y = "Yield_tn",
      x = "Treatment",
      cellsize = 9,
      nmin_cell = 4,
      n_p = 2000,
      n_s = 200,
      alpha = 0.05,
      shift = 0,
      alpha_bonferroni = FALSE,
      crs = NULL) # en la salida corregir que diga MI en lugar IM

```

The results show that out of the total number of cells selected (554), the ESS 
was 97, with a magnitude of autocorrelation (Rho) of 0.65. The fertilization
treatment showed significant differences compared to the control (p < 0.05).

The average yield of the fertilized strip was 5.28 $t~ha^{-1}$, while that of the
control was 4.76 $t~ha^{-1}$.


## References
Córdoba M., Paccioretti P.,Balzarini M. 2024. A new method to compare treatments
in unreplicated on-farm experimentation. Precis. Agric. (under review). 

Vega A., Córdoba M., Balzarini M. 2019. Protocol for automating error removal
from yield maps. Precis. Agric. 20: 1030–1044. 
https://doi.org/10.1007/s11119-018-09632-8
