---
title: "Getting started with ofemeantest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ofemeantest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

`ofemeantest` implements a cell-based, permutation-based protocol to
compare treatments in **unreplicated on-farm experiments** (OFE) where
strips of management are laid out side-by-side over a field and yield is
recorded densely (e.g., from a yield monitor).

The package handles three things:

1. **Build a regular grid** over the trial and aggregate the dense point
   data into cell-level medians, keeping only cells that fall inside a
   single treatment strip and contain enough observations.
2. **Estimate the effective sample size (ESS)** from the spatial
   autocorrelation of the residuals of a one-way ANOVA on cell medians,
   to avoid pseudo-replication from spatially correlated data.
3. **Test treatment differences** with a permutational ANOVA, repeated
   over many random subsamples of size ESS, and report the median
   *p*-value as the field-specific test result.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pkg_install("PPaccioretti/ofemeantest")
```

## A worked example

The bundled `ofe_f2` dataset comes from a corn (Zea mays L.) trial where
a single fertilized strip (~2.2 ha) was compared against an adjacent
control strip on the same field. Raw yield-monitor data were cleaned
following Vega et al. (2019).

```{r load}
library(ofemeantest)
data("ofe_f2")

head(ofe_f2)
```

### Quick run

Call `ofemt()` and let it build the grid internally:

```{r quick, message = FALSE, warning = FALSE}
res <- ofemt(
  data = ofe_f2,
  y = "Yield_tn",
  x = "Treatment",
  cellsize = 9,
  min_per_cell = 4,
  n_p = 2000,
  n_s = 200,
  alpha = 0.05
)
res
```

Key fields in the printed result:

- **Selected cells**: cells retained after filtering by single
  treatment and `min_per_cell`.
- **n / ESS**: nominal sample size and effective sample size given
  the spatial autocorrelation of the residuals.
- **Rho / Moran's I**: spatial autocorrelation indicators.
- **Means comparison**: per-treatment median yield (in cells) and
  compact letter display from the multiple-comparison procedure.
- **ANOVA permutation test**: median *p*-value per pairwise comparison.

### Step-by-step

For finer control — inspecting the grid, tweaking the cell size, or
reusing the same selection in several downstream analyses — build the
grid explicitly with `make_ofe_grid()` and pass it to `ofemt()`.

```{r stepwise, message = FALSE, warning = FALSE}
g <- make_ofe_grid(
  data = ofe_f2,
  x = "Treatment",
  cellsize = 9,
  min_per_cell = 4
)
names(g)


plot_grid_selection(g, data = ofe_f2)

res2 <- ofemt(
  data = ofe_f2,
  y = "Yield_tn",
  x = "Treatment",
  grid = g,
  n_p = 2000,
  n_s = 200
)
```

Both calls produce the same numeric output when the grid is built with
matching arguments — `ofemt(grid = NULL, ...)` is equivalent to
`ofemt(grid = make_ofe_grid(...))` under the hood.

### Tuning the grid

`plot_grid_selection()` draws the three layers on one set of axes: the
full grid in grey, the cells that survived the filters shaded in blue,
and the observations as points. Overlaying the points on the cell
boundaries is what makes the geometric arguments legible — a cell is
dropped either because it straddles two treatments or because too few
points landed inside it, and both are visible at a glance.

The title of each plot repeats the parameters used, so successive calls
can be compared directly:

```{r tuning, eval = FALSE}
op <- par(mfrow = c(1, 2))
plot_grid_selection(
  make_ofe_grid(ofe_f2, x = "Treatment", cellsize = 9, min_per_cell = 4),
  data = ofe_f2
)
# Shift the origin by half a cell and rotate to follow the strips
plot_grid_selection(
  make_ofe_grid(
    ofe_f2,
    x = "Treatment",
    cellsize = 9,
    min_per_cell = 4,
    shift = c(4.5, 4.5),
    angle_deg = 10,
    buffer = 5
  ),
  data = ofe_f2
)
par(op)
```

### Keeping the geometries in the result

By default `ofemt()` returns tables only. Set `keep_components` to embed
the spatial objects in the result, which lets you plot the analysis that
actually ran rather than rebuilding the grid by hand:

| `keep_components` | Adds to the result | `plot()` shows |
|---|---|---|
| `"none"` (default) | nothing | *(errors — no geometries)* |
| `"light"` | `grid`, the full `ofe_grid` (`grid_all`, `grid_sel`, `cell_stats`, `params`) | grid + selected cells |
| `"full"` | `"light"` plus `cell_medians` (one point per selected cell, with its median response and ANOVA residual) and `points_joined` (every observation with its `CellID`) | grid + selected cells + points |

Cost scales accordingly: `"light"` stores one polygon per grid cell,
`"full"` adds one row per observation.

```{r keep, eval = FALSE}
res_full <- ofemt(
  ofe_f2,
  y = "Yield_tn",
  x = "Treatment",
  cellsize = 9,
  min_per_cell = 4,
  keep_components = "full"
)

# No further arguments needed — every layer comes from the object itself
plot(res_full)

# The per-cell medians and residuals that fed the spatial diagnostics
head(res_full$cell_medians)
```

### Reproducibility

`ofemt()` exposes a `seed` argument that controls the random subsampling
inside the permutation runs. The default is `seed = 7L`, so two calls
with the same inputs return identical p-values:

```{r seed, eval = FALSE}
identical(
  ofemt(
    ofe_f2,
    y = "Yield_tn",
    x = "Treatment",
    cellsize = 9,
    min_per_cell = 4,
    seed = 7L
  ),
  ofemt(
    ofe_f2,
    y = "Yield_tn",
    x = "Treatment",
    cellsize = 9,
    min_per_cell = 4,
    seed = 7L
  )
)
#> TRUE
```

Pass `seed = NULL` to let results vary across runs (e.g., when
exploring sensitivity to the random draws).

### Inspecting the permutation distribution

Each call retains the per-run *p*-values in `res$perm_runs` so you can
sanity-check that the median *p*-value is not an artefact of a long
tail. `perm_runs` has one row per comparison per sampling run, with
columns `Comparison`, `p_value` (that run's permutation *p*-value),
`p_adj` (the same value after adjusting for multiplicity *within* the
run) and `run`.

`plot_pvalue_hist()` draws that distribution with two reference lines
per panel: a solid line at the median — the value reported in the
`ANOVA permutation test` table — and a dashed line at `alpha`.

```{r hist, eval = FALSE}
# Requires the optional 'ggplot2' package.
plot_pvalue_hist(res)
```

When the analysis was run with a multiplicity adjustment, the histogram
shows the *adjusted* values by default, so the median line and the
reported `p_adj` always refer to the same quantity. Use
`which = "raw"` to see the unadjusted distribution instead:

```{r hist_adj, eval = FALSE}
res_bonf <- ofemt(
  ofe_f2,
  y = "Yield_tn",
  x = "Treatment",
  cellsize = 9,
  min_per_cell = 4,
  p_adjust_method = "bonferroni"
)

plot_pvalue_hist(res_bonf)                 # adjusted
plot_pvalue_hist(res_bonf, which = "raw")  # unadjusted
```

## Where to go next

- `vignette("methodology")` — technical description of the protocol
  (effective sample size, permutational ANOVA, multiplicity adjustment).
- `?ofemt` — full argument reference for the main function.
- `?make_ofe_grid` — details on grid construction and selection.

## References

- Córdoba M., Paccioretti P., Balzarini M. (in review). A new method
  to compare treatments in unreplicated on-farm experimentation.
- Griffith, D.A. (2005). Effective geographic sample size in the
  presence of spatial autocorrelation. *Annals of the Association of
  American Geographers* 95(4): 740–760.
- Vega A., Córdoba M., Balzarini M. (2019). Protocol for automating
  error removal from yield maps. *Precision Agriculture* 20: 1030–1044.
  <https://doi.org/10.1007/s11119-018-09632-8>
