---
title: "Getting started with ofemeantest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ofemeantest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

`ofemeantest` implements a cell-based, permutation-based protocol to
compare treatments in **unreplicated on-farm experiments** (OFE) where
strips of management are laid out side-by-side over a field and yield is
recorded densely (e.g., from a yield monitor).

The package handles three things:

1. **Build a regular grid** over the trial and aggregate the dense point
   data into cell-level medians, keeping only cells that fall inside a
   single treatment strip and contain enough observations.
2. **Estimate the effective sample size (ESS)** from the spatial
   autocorrelation of the residuals of a one-way ANOVA on cell medians,
   to avoid pseudo-replication from spatially correlated data.
3. **Test treatment differences** with a permutational ANOVA, repeated
   over many random subsamples of size ESS, and report the median
   *p*-value as the field-specific test result.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pkg_install("PPaccioretti/ofemeantest")
```

## A worked example

The bundled `ofe_f2` dataset comes from a corn (Zea mays L.) trial where
a single fertilized strip (~2.2 ha) was compared against an adjacent
control strip on the same field. Raw yield-monitor data were cleaned
following Vega et al. (2019).

```{r load}
library(ofemeantest)
data("ofe_f2")

ofe_f2
```

### Quick run

Call `ofemt()` and let it build the grid internally:

```{r quick, message = FALSE, warning = FALSE}
res <- ofemt(
  data         = ofe_f2,
  y            = "Yield_tn",
  x            = "Treatment",
  cellsize     = 9,
  min_per_cell = 4,
  n_p          = 2000,
  n_s          = 200,
  alpha        = 0.05
)
res
```

Key fields in the printed result:

- **Selected cells**: cells retained after filtering by single
  treatment and `min_per_cell`.
- **n / ESS**: nominal sample size and effective sample size given
  the spatial autocorrelation of the residuals.
- **Rho / Moran's I**: spatial autocorrelation indicators.
- **Means comparison**: per-treatment median yield (in cells) and
  compact letter display from the multiple-comparison procedure.
- **ANOVA permutation test**: median *p*-value per pairwise comparison.

### Step-by-step

For finer control — inspecting the grid, tweaking the cell size, or
reusing the same selection in several downstream analyses — build the
grid explicitly with `make_ofe_grid()` and pass it to `ofemt()`.

```{r stepwise, message = FALSE, warning = FALSE}
g <- make_ofe_grid(
  data         = ofe_f2,
  x            = "Treatment",
  cellsize     = 9,
  min_per_cell = 4
)
g

plot_grid_selection(g)

res2 <- ofemt(
  data = ofe_f2,
  y    = "Yield_tn",
  x    = "Treatment",
  grid = g,
  n_p  = 2000,
  n_s  = 200
)
```

Both calls produce the same numeric output when the grid is built with
matching arguments — `ofemt(grid = NULL, ...)` is equivalent to
`ofemt(grid = make_ofe_grid(...))` under the hood.

### Reproducibility

`ofemt()` exposes a `seed` argument that controls the random subsampling
inside the permutation runs. The default is `seed = 7L`, so two calls
with the same inputs return identical p-values:

```{r seed, eval = FALSE}
identical(
  ofemt(ofe_f2, y = "Yield_tn", x = "Treatment", cellsize = 9, min_per_cell = 4, seed = 7L),
  ofemt(ofe_f2, y = "Yield_tn", x = "Treatment", cellsize = 9, min_per_cell = 4, seed = 7L)
)
#> TRUE
```

Pass `seed = NULL` to let results vary across runs (e.g., when
exploring sensitivity to the random draws).

### Inspecting the permutation distribution

Each call retains the per-run *p*-values in `res$perm_runs` so you can
sanity-check that the median *p*-value is not an artefact of a long
tail:

```{r hist, eval = FALSE}
# Requires the optional 'ggplot2' package.
plot_pvalue_hist(res)
```

## Where to go next

- `vignette("methodology")` — technical description of the protocol
  (effective sample size, permutational ANOVA, multiplicity adjustment).
- `?ofemt` — full argument reference for the main function.
- `?make_ofe_grid` — details on grid construction and selection.

## References

- Córdoba M., Paccioretti P., Balzarini M. (in review). A new method
  to compare treatments in unreplicated on-farm experimentation.
- Griffith, D.A. (2005). Effective geographic sample size in the
  presence of spatial autocorrelation. *Annals of the Association of
  American Geographers* 95(4): 740–760.
- Vega A., Córdoba M., Balzarini M. (2019). Protocol for automating
  error removal from yield maps. *Precision Agriculture* 20: 1030–1044.
  <https://doi.org/10.1007/s11119-018-09632-8>
